---
title: "Housing Tenure in Atlanta metro jurisdictions"
author: "Taylor Hafley"
date: "5/22/2021"
output: 
  pdf_document: default
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(echo = FALSE, dev = "cairo_pdf", warning=FALSE, message=FALSE,knitr.table.format = "latex")
# fig.width=5.6, fig.height=4,

# load libraries
library(tidyverse)
library(tidycensus)
library(sf)
library(tmap)
library(tigris)
options(tigris_use_cache = TRUE)
options(tigris_class = 'sf')
library(hrbrthemes)
library(ggplot2)
library(kableExtra)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
library(areal)
library(viridis)
th_api_acs <- '45544f0d114cfaa037a5566745d18bb8d4778cfa'

####-----------------------------------------------------------------------
## Retrieve data from census bureau using tigris package
## counties boundary
## roads (for cartographic representation)
####-----------------------------------------------------------------------

# create list of 19 counties where Invitation Homes owns housing
metro_counties <- c("Gwinnett", "Barrow", "Bartow", "Carroll", "Cherokee","Clayton","Coweta", 
                    "DeKalb", "Douglas", "Fayette","Forsyth", "Gwinnett", 
                    "Hall","Henry", "Newton", "Paulding", "Rockdale", "Walton", "Cobb", "Fulton"
                    )


# bring in county boundaries where INVH is active via tigris
counties_sf <- counties("GA") %>%      # Georgia counties
  filter(NAME %in% metro_counties)     # filter to metro

# project
counties_sf <- counties_sf %>%
  st_transform(26916)

# primary roads for study area
rd <- primary_roads()
rd <- rd %>%
  st_transform(26916)
rd <- st_crop(rd,counties_sf) #crop roads to study area


###-------------------------------------------------
# read in georgia municipalites from Atlanta Regional Commission open data
ga_cities_api = read_sf("https://opendata.arcgis.com/datasets/34520575dfc34b8cac783caff702b8cc_58.geojson")
st_is_valid(ga_cities_api)

# correct geometry errors
ga_cities_api <- st_make_valid(ga_cities_api)
st_is_valid(ga_cities_api)
# good to go. all geoms fixed.

# project
ga_cities_api <- ga_cities_api %>%
  st_transform(26916) #EPSG 26916 is projected: NAD83 / UTM zone 16N # EPSG:26916 Projected coordinate system for North America

# cut to 20 metropolitan counties, and delete municipalities in Spalding County, 
# Invitation Homes does not own property in Spalding County
spalding_munis <- c("Griffin", "Sunny Side", 'Orchard Hill')

atl_munis <- ga_cities_api %>%
  filter(County20 == "yes") %>% #per metadata, County20 == Within ARC 20 County region
  filter(!(Name %in% spalding_munis)) # drop Spalding

# 119 municipalities in 19 county study area


####-----------------------------------------------------------------------
## Read in Invitation Homes geolocated data
## these data were retrieved from public facing tax parcel data
## cleaned, geocoded using google api, and joined together in a separate project
## I would prefer to keep this data private as this is unpublished research
## If you need it, I will make it available on github so that the script will run
## 11,633 addresses
####-----------------------------------------------------------------------

# read in geocoded address level data for INVH properties
ih_dots <- read_sf("../../invh/metro/data/ih_points_ct.geojson") # data I would prefer not to share. 

ih_dots <- ih_dots %>%
  st_transform(26916) #EPSG 26916 is projected: NAD83 / UTM zone 16N # EPSG:26916 Projected coordinate system for North America

####-----------------------------------------------------------------------
## test data

# set tmap mode
tmap_mode('plot')
# test map with municipal boundaries, county boundaries, and geolocation data
tm_shape(atl_munis) + 
  tm_borders("red", lwd = .9) +
  tm_shape(ih_dots) +
  tm_dots("gray",size = .01) +
  tm_shape(counties_sf) +
  tm_borders(lwd = 2) +
  tm_shape(rd) +
  tm_lines()
# good to go!

# clean up this subsection
rm(ga_cities_api)
```

## Executive Summary

## Introduction
### !uestions
1. what county has the most non-incorporated SFRs? answered, Gwinnett holds big change is Fulton 
2. at the municipality, what is IH's share of the SF stock and SFR stock?


### How do we respond to reviewers who might say there are low percentages?
1. this is a single entity, which gives them considerable ownership of the housing stock
available in this municipality. Two other publicly-listed SFRs operating in metropolitan Atlanta, own, in tandem, about the number of houses equivalent to Invitation Homes.

### Why are these important questions?
1. these findings are relevent to public and elected officials of these jurisdictions as they
discuss issues related to zoning, housing affordability, housing supply, and homeownership


## Scope of Work
This document does four things
1. Map the municipal geography of Invitation Homes
2. Show how municipalities compare to non-incorporated spaces
3. Show the presence in relation to the municipal housing stock
4. Analyze the demographics of the municipalities where Invitation Homes is most concentrated

### TOC
### Introduction
### Methods and Data Collection
### Findings
### Discussion

```{r cars}

####-----------------------------------------------------------------------
## Identify if geolocated address data are in a municipality
## Count how many locations are in each municipality
## st_intersection help: https://github.com/r-spatial/sf/issues/347
####-----------------------------------------------------------------------

## This identifies and returns the municipality each address is in
muni_intersection <- st_intersection(x = atl_munis, y = ih_dots)

####-----------------------------------------------------------------------
## We could create a file from this.
## create point file (i.e. 'metro' that includes GEOID for each location)
## write_csv(muni_intersection, "ih_dots_muni.csv")
####------------------------------------------------------------------------

# # Check to see if any address was assigned to more than one municipality
# muni_intersection %>%
#   as_tibble() %>%
#   select(-(geometry)) %>%   # this line makes the code run faster
#   group_by(Parcel.ID) %>%   # parcel uniqe to each 
#   summarise(N = n()) %>%
#   filter(!(N == 1))         # return any parcels identified more than once

# There are two records with two parcel IDs. This is because the geolocation overlapped with two city boundaries
# these are the two parcels:
# 1 09F020200131418 # Fairborn and South Fulton -- Fairborn, per Fulton County website: https://iaspublicaccess.fultoncountyga.gov/search/CommonSearch.aspx?mode=PARID
# 2 09F280001111122 # Union City and South Fulton -- Union City, per Fulton County website: https://iaspublicaccess.fultoncountyga.gov/search/CommonSearch.aspx?mode=PARID

# to correct, remove the two rows/parcels identified as in 'South Fulton'
# incorp_dots is all addresses within incorporated municipalities
incorp_dots <- muni_intersection %>%
  as_tibble() %>%
  select(-(geometry)) %>%
  filter(!(Parcel.ID == "09F020200131418" & Name == 'South Fulton' | Parcel.ID == "09F280001111122" & Name == 'South Fulton'))

# group all addresses by (muni) name, then sum, and create new tbl 
munis_n <- incorp_dots %>% 
  group_by(Name) %>%  # municipality name
  summarise(ihsfrs = n()) # ihsfrs = the number of addresses/points in each municipality

# INVH owns housing in 83 of the 119 municipalities

# # how many properties are in the municipalities? 
# munis_n %>%
#   summarize(total = sum(ihsfrs)
#   )
# #3382, correct.

# arrange municipalities, in descending order, by number of properties Ivntiation Homes owns.
# munis_n %>%
#   arrange(desc(ihprops))

# join ihprops count to atl_munis sf in order to create a 
# file with all 119 municipalities and IH totals
munis2 <- left_join(atl_munis, munis_n, by = "Name")

# replace NAs (munis where INVH owns 0 SFRs) with 0
munis2 <- munis2 %>%
  replace_na(list(ihsfrs = 0)) %>%
  select(OBJECTID, Name, ihsfrs, Acres, Sq_Miles) # clean up sf. don't need County20, County10, GlobalID, and 'last_edited'

# # looking at data
# munis2 %>%
#   filter(ihsfrs > 20) %>%
#   select(Name, ihsfrs) %>%
#   arrange(desc(ihsfrs)) %>%
#   print(n = 30)
# 
# # testing
# munis2 %>%
#   summarise(total = sum(ihsfrs))
# 
# # 3382/11632
# # 29.0% of ih market is in incorporated spaces
# # 7/10 in unincorporated space is pretty striking. 

# map it
tm_shape(munis2) +
  tm_polygons("ihsfrs", palette = 'viridis', style = 'jenks', lwd = 0.1) +
  tm_shape(counties_sf) +
  tm_borders()

# munis2

# clean up this section
rm(munis_n, atl_munis)
# muni_intersection?


```

### An Invitaion Homes Primer
## Data Collection
## Municipalities
Only ~30% of Invitation Homes' portfolio is within municipalities

### The counties
~70% in unincorporated areas. But, the municipalities are not evenly distributed, so I analyze by each county to see how the totals compare between incorporated and unincorporaed spaces. 


```{r pressure, echo=FALSE}
####-----------------------------------------------------------------------
## county analysis. In the next series of steps, I calculate the number of IH-owned
## SFRs in each county, within a municipality in each county, and outside of a municipality
## in each county
## 
####-----------------------------------------------------------------------
# create county tbl calculating sum of ih owned sfrs in incorporated areas in each county
# based on addresses identified in line 213
incorp_county <- incorp_dots %>%
  mutate(county_fips = str_sub(GEOID,1,5)) %>%  # create county fips column from tract
  group_by(county_fips) %>%
  summarise(
    incorp_ihsfrs = sum(N = n()) # unincorp_ihsfrs == sum of ih owned sfrs in unincorporated areas
  )

# incorp_county %>%
#   summarise(N = sum(incorp_ihsfrs))
# #3,382 == correct

####-----------------------------------------------------------------------
# create sf of addresses outside of municipalities
# using anti-join of all address and incorporated address
unincorp_dots <- anti_join(ih_dots, incorp_dots, by = 'Parcel.ID')

# create county tbl calculating sum of ih owned sfrs in unincorporated areas in each county
# based on addresses identified in line 213
unincorp_county <- unincorp_dots %>%
  as_tibble() %>%
  select(-(geometry)) %>%
  mutate(county_fips = str_sub(GEOID,1,5)) %>%  # create county fips column from tract
  group_by(county_fips) %>%
  summarise(
    unincorp_ihsfrs = sum(N = n()) # unincorp_ihsfrs == sum of ih owned sfrs in unincorporated areas
  )

# unincorp_county %>%
#   summarise(N = sum(unincorp_ihsfrs))
# 8251(unincorporated)+ 3382(incorp_dots) == 11633
# all good.

####-----------------------------------------------------------------------
# create county tbl calculating sum of ih owned sfrs in each county
# based on all addresses
total_county <- ih_dots %>%
  as_tibble() %>%
  select(-(geometry)) %>%
  mutate(county_fips = str_sub(GEOID,1,5)) %>%
  group_by(county_fips) %>%
  summarise(
    county_ihsfrs = sum(N = n())
  )

# join 
ih_counties <- left_join(counties_sf, total_county, by = c("GEOID" = "county_fips")) %>%
  left_join(., unincorp_county, by = c("GEOID" = "county_fips")) %>%
  left_join(., incorp_county, by = c("GEOID" = "county_fips")
            )

# check
# ih_counties %>%
#   summarize(check = county_ihsfrs -(incorp_ihsfrs + unincorp_ihsfrs)
#             ) %>%
#   print(n = 19)

# project
ih_counties<- ih_counties %>%
  st_transform(26916)

# compare distribution across counties
# tm_shape(ih_counties) +
#   tm_polygons(c("unincorp_ih_sfrs","incorp_ih_sfrs","ihprops"), style = 'jenks')

# create shorter county name in preparation of figures
ih_counties <- ih_counties %>%
  mutate(short_name = '',
         short_name = sub(" .*", "", NAME)) %>%
  select(short_name, everything()) 
```


```{r Census Data}
####-----------------------------------------------------------------------
## read in census data. this section of code was adapted from :
## https://github.com/jshannon75/awp_tidycensus/blob/master/ACS_walkthrough.md
####-----------------------------------------------------------------------

# existing .csv of variables of interest to function as a lookup table
metadata <- read_csv("acs_vars_metadata.csv")

# variable codes of interest
acs_vars <- metadata$variable

# using tidycensus, pull in census tract data from 2014-18 ACS
atl_ct <- get_acs(geography = "tract", variables = acs_vars, state = "GA",
                  county = metro_counties, year = 2018,
                  geometry = FALSE, key = th_api_acs) %>% # no geometry to make data modifications for MOE run smoothly. Will join to spatial in line 328
  mutate(GEOID=as.character(GEOID)
         )

atl_ct2 <- atl_ct %>%
  left_join(metadata) %>%
  group_by(GEOID,var_group) %>% #var_name or var_group? #shannon says group, i think name
  summarise(
    est=sum(estimate),
    moe=round(sqrt(sum(moe^2)),0)
  )

# make wide to join with sf and prepare for areal package
atl_wide <- atl_ct2 %>%
  pivot_wider(
    names_from = var_group,
    values_from = c(est,moe),
    names_glue = "{var_group}_{.value}"
  )

# get tract geographies
atl_sf <- get_acs(geography = "tract", variables = "B19001_001", state = "GA",
                  county = metro_counties, year = 2018,
                  geometry = TRUE, key = th_api_acs) %>%
  mutate(GEOID=as.character(GEOID)) %>%
  select(-c("variable", "estimate", "moe")) # don't need the variable

# join sf to correct census data
atl_wide <- left_join(atl_sf,atl_wide)

# project
atl_wide <- atl_wide %>%
  st_transform(26916)

# remove missing data in ATL metro
atl_wide <- atl_wide %>%
  filter(!(GEOID %in% c(13089023115, 13121006801, 13063980000,13089980000,13121980000)))

# clean up section
rm(atl_ct,atl_ct2,atl_sf)
```




```{r Interpolation}
####-----------------------------------------------------------------------------
# Area  Interpolation with 'areal' package
####-----------------------------------------------------------------------------

# # area interopolation
# # take tract geography and take some of listed variables for each census tract to municipality
muni_tenure <- aw_interpolate(munis2, tid = OBJECTID, source = atl_wide, sid = "GEOID",
                              weight = "sum", output = "sf", extensive = c("rntocc_sf_est","tot_sf_est"))


# create ratios of Invitation Homes' ownership to normalize based on single-family rentals and single-family units
muni_tenure <- muni_tenure %>%
  mutate(pct_ih = round(ihsfrs/tot_sf_est*100000,1),
         pct_ihsfr = round(ihsfrs/rntocc_sf_est*100000,1))

####-----------------------------------------------------------------------
## figures
####-----------------------------------------------------------------------

tm_shape(muni_tenure) +
  tm_polygons("pct_ihsfr", palette = 'viridis', style = 'jenks', lwd = 0.1) +
  tm_shape(counties_sf) +
  tm_borders()

tm_shape(muni_tenure) +
  tm_polygons("pct_ih", palette = 'viridis', style = 'jenks', lwd = 0.1, id = 'Name') +
  tm_shape(counties_sf) +
  tm_borders()

library(ggrepel)
ggplot(data = muni_tenure, mapping = aes(x = ihsfrs, y = pct_ihsfr)) +
  geom_point() +
  geom_smooth() +
  geom_text_repel(data = subset(muni_tenure, ihsfrs > 150 | pct_ihsfr >6000),
                  mapping = aes(label = Name)) +
  # arrow = arrow(length = unit(0.02, "npc")),
  # box.padding = 1) +
  theme_ipsum()


# muni_tenure %>%
#   select(OBJECTID, Name, ihsfrs, pct_ihsfr, pct_ih) %>%
#   arrange(desc(pct_ihsfr)) %>%
#   print(n=15)
```

\newpage
## Findings
At this point, I have done all of the interpolation, data cleaning and wrangling

```{r final}
quantile(munis2$ihsfrs, probs = seq(0, 1, by= 0.1)) # decile
summary(munis2$ihsfrs)
# 70th percentile is 23.8; median is 5 SFRs
# max is 418

# create figure of municipalities where Invitation Homes owns >50 single-family units.
# 51.4 is 80th percentile; top 25 municipalities
munis2 %>%
  filter(ihsfrs >50) %>%
  ggplot(munis2, mapping = aes(x = reorder(Name,ihsfrs), y = ihsfrs, label = ihsfrs)) +
  geom_col(fill = "#41912E", position = 'dodge') +
  labs(x = NULL, y = "Total SFRs",
       title = "Invitation Homes SFRs, by Municipality") +
  geom_text(size = 2.5, position = position_stack(vjust = 0.5)) +
  theme_ipsum(axis_text_size = 9) +
  coord_flip()

# create tbl of key variables: ihprops, all sfs, all sfrs
munis2 %>%
  as_tibble() %>%
  select(-(geometry)) %>%
  select(Name, ihsfrs) %>%
  arrange(Name) %>%
  kable(caption = "Inviitation Homes, by SFR", #, escape = F, booktabs =T,
        # format.args = list(big.mark = ",", scientic = FALSE),
        col.names = c("Municipality",
                      "INVH SFRs"))

# ggplot(ih_counties, aes(x = reorder(short_name,unincorp_ihsfrs), y = unincorp_ihsfrs, label = unincorp_ihsfrs)) +
#   geom_col(fill = "#AADE55", position = 'dodge') +
#   labs(x = NULL, y = "Total SFRs",
#        title = "Invitation Homes SFRs, by County, Unincorporated Area") +
#   geom_text(size = 2.5, position = position_stack(vjust = 0.5)) +
#   scale_y_continuous(labels = scales::comma) +
#   theme_ipsum(axis_text_size = 9) +
#   coord_flip()

ih_counties_long <- ih_counties %>%
  pivot_longer(cols = c(unincorp_ihsfrs,incorp_ihsfrs), names_to = "var", values_to = 'count')

ggplot(ih_counties_long, aes(x = reorder(short_name,count), y = count, fill = var, label = count )) +
  geom_col(position = 'stack') +
  scale_fill_manual(values=c("#AADE55","#41912E"), labels = c("Unincorporated SFRs", "Incorporated SFRs")) +
  labs(x = NULL, y = "SFRs",
       title = "Atlanta's Largest Land owner concentrated outside Municipalities",
       subtitle = "Invitation Homes' SFRs, by jurisdiction",
       fill = "") +
  # geom_text(size = 2.5, position = position_stack(vjust = 0.5)) +
  # theme(axis.text.x=element_text(angle=60, hjust=1)) +
  theme_ipsum(axis_text_size = 9) +
  theme(legend.position  = "bottom",
        legend.text = element_text(size = rel(0.9)),
        legend.key = element_rect(size = 1)) +
  coord_flip()
```

